Split-Apply-Combine with Dynamic Grouping

### 2023-02-21 paper/article.Rnw
wip

\begin{algorithm}[H]
\caption{Split-Apply-Combine}
\label{alg:sac}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, and a partition
       $f:U\onto A$.}
\Output{$R$: the value of $\phi$ for every part of $U$ as a set of pairs
        $(a,x)\in A\times X$. }
$R = \{\}$\;
\For{$a\in A$}{
  $d=f^*(\{a\})$\tcp*{get subset of $U$}
  $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
}
\end{algorithm}
%
In this algorithm the output is collected in a set $R$ containing pairs from
$A\times X$: one pair for each element of $A$. Incidentely, the algorithm can
be summarized even shorter in this notation as $R=\cup_{a\in A}(a,(\phi\circ
f^*)(\{a\}))$, where $\circ$ denotes function composition.  Since $f^*$ always
receives a singleton set, its definition as a function of the powerset of the
group labels $A$ is too general for this algorithm.  However, this
generalization turns out to be necessary when generalizing the algorithm to the
case of dynamic grouping.

@@

\subsection{Split-apply-combine with collapsing groups}
\label{sect:saccg}
The goal of the algorithm is to compute a value for each part of a dataset,
possibly using values external to the part. The input of the algorithm consists
again of a finite set $U$ and an aggregation function $\phi$ that takes a
subset of $U$ and returns a value in some domain $X$.  Compared to
Algorithm~\ref{alg:sac} two different inputs are needed. First, a function must
be defined that that checks whether a given subset $d$ of $U$ is suitable for
computing $\phi(d)$.  We will call this function $\beta: 2^U\to \mathbb{B}$,
where $\mathbb{B}=\{\texttt{True},\texttt{False}\}$.  Typical tests are
checking whether there are sufficient records available, or whether certain
variables have a low enough fraction of missing values. Second, we need a
\emph{collapsing scheme}
\begin{equation}
U\xonto{f_0}A_0\xonto{f_1}A_1\xonto{f_2}\cdots\xonto{f_n}A_n.
\label{eq:collapsingsequence}
\end{equation}
A collapsing scheme is a sequence of partitions where each $f_i:A_{i-1}\to A_i$
partitions $A_{i-1}$ into $|A_i|$ groups.

Denote with $F_k:A_0\to A_k$, the function that  accepts a label in $A_0$ and
returns the corresponding label in $A_k$. In other words, $F_k$ is the
composition $f_k\circ f_{k-1}\circ\cdots \circ f_1$. Similarly we define the
pullback along $F_k$ as $F_k^* = f_1^*\circ f_2^*\circ\cdots\circ f_k^*$.  This
function accepts a set of labels in $A_k$ and returns all the labels in $A_0$
that are mapped to those labels via the collapsing sequence of
Equation~\eqref{eq:collapsingsequence}.  With this notation we can define the
Algorithm for aggregation with dynamic grouping as follows.

\begin{algorithm}[H]
\caption{Split-Apply-Combine with Collapsing Groups}
\label{alg:saccg}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, a test function $\beta: 2^U\to \mathbb{B}$,
      and a collapsing sequence $U\xonto{f_0}A_0\xonto{f_1}\cdots \xonto{f_n}  A_n$}

\Output{$R$: the value of $\phi$ for every part of $U$, for which a suitable
collapsing group can be found, as a set of triples $(a,k,x)\in A_0\times
\underline{n}\times X $ where $\underline{n}=\{0,1,\ldots,n\}$.}

$R = \{\}$\;
\For{$a\in A_0$}{
  $i=0$ \tcp*{Initiate collapse level}
  $d = f_0^*(\{a\})$ \tcp*{Get subset of $U$}
  \While{$i<n \land \lnot\beta(d)$}{ \label{line:while}
    $i = i+1$ \tcp*{Increase collapse level} 
    $d = (f_0^*\circ F_i^*\circ F_i)(a)$ \tcp*{Collapse and get subset}
  }\label{line:endwhile}
  \If{$i<n \lor \beta(d)$}{ \label{line:cond}
    $R = R\cup \{(a, i,\phi(d))\}$\;
  } 
}

@@

In this algorithm the collapsing level $i$ is increased until the test is
passed or the maximum collapsing level $n$ is reached
(Lines~\ref{line:while}-\ref{line:endwhile}) %Since the eventual level of
collapsing is determined dynamically by data %circumstances in $U$, the
algorithm also reports the collapsing level $i$ used.  The condition in
Line~\ref{line:cond} ensures that if no suitable dataset is found after the
whole collapsing sequence has been executed, then no answer is returned. This
means that in contrast with Algorithm~\ref{alg:sac} there is no guarantee that
a value for each member of $A_0$ will be found. For each member of $A_0$ where
an aggregate is computed, there is a triple $(a,i,\phi(d))$, where $a\in A_0$
is the label to which the value $\phi(d)$ pertains, and $i$ is the number of
collapses applied to reach a suitable dataset.

\subsection{Performance, memory, parallelization}
Both Algorithms~\ref{alg:sac} and \ref{alg:saccg} need only $\mathcal{O}(|U|)$
memory. The time complexity of Algorithm~\ref{alg:sac} is $\mathcal{O}(|A||U|)$
($|A|$ the number of groups occurring in the data, and $|U|$ the number of
records) but the actual running time is also determined by the form of the
aggregating function $\phi$.  For simple aggregating functions such as
summation or counting, the output can be computed with a single pass through
$|U|$. \proglang{R} packages such as \pkg{collapse} and \pkg{data.table} have
implemented specialized low level (\proglang{C}/\proglang{C++}) code for fast
groupwise aggregation for some commonly occurring aggregating functions.
Generic parallelization of Algorithm~\ref{alg:sac} can be done by parallelizing
over the grouping values $a\in A$: each worker node gets access to a set of
records corresponding to a single group and computes the local aggregate.

Algorithm~\ref{alg:saccg} has time complexity of $\mathcal{O}((n+1)|A_0||U|)$,
Namely, in the worst case, for each $a\in A_0$ we need to collapse to the
maximum level $A_n$. This means that for each $a\in A_0$ we need to collapse
$n$ times, plus the initial attempt of computing aggregate without collapsing.
The fact that we cannot generically exclude records from a calculation means
that a generic parallelization scheme would need to ensure that each worker
node has access to the full dataset $|U|$. There are two types of options to do
this: either multliple workers gain read-only access to $U$ or each worker gets
a copy of $U$. In the first case the workers may suffer from colliding
read-operations which hampers linear performance improvement as a function of
the number of workers. In the second case, one needs as many copies of the data
as there are threads. It is possible to compute beforehand whether the dataset
can be split into subsets that can be treated independently by determining all
the collapsed subsets. However, this is almost as expensive as running the
algorithm completely since it only skips the calculation of aggregates after
collapsing.

Summarizing, we see that split-apply-combine with collapsing groups has a
higher worst-case complexity than regular split-apply-combine. Moreover,
generic parallelization needs to either consider a specific use case (e.g.
large or small datasets) or require user input that guides the choice of
parallelization scheme.

>>>>

\begin{algorithm}[H]
\caption{Split-Apply-Combine}
\label{alg:sac}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, and a partition
       $f:U\onto A$.}
\Output{$R$: the value of $\phi$ for every part of $U$ as a set of pairs
        $(a,x)\in A\times X$. }
$R = \{\}$\;
\For{$a\in A$}{
  $d=f^*(\{a\})$\tcp*{get subset of $U$}
  $R = R\cup \{(a,\phi(d))\}$\tcp*{aggregate and add to result}
}
\end{algorithm}
%
In this algorithm the output is collected in a set $R$ containing pairs from
$A\times X$: one pair for each element of $A$. Incidentely, the algorithm can
be summarized even shorter in this notation as $R=\cup_{a\in A}\{(a,(\phi\circ
f^*)(\{a\}))\}$, where $\circ$ denotes function composition.  Since $f^*$
always receives a singleton set, its definition as a function of the powerset
of the group labels $A$ is too general for this algorithm.  However, this
generalization turns out to be necessary when generalizing the algorithm to the
case of dynamic grouping.

@@

\subsection{Split-apply-combine with collapsing groups}
\label{sect:saccg}
The goal of the algorithm is to compute a value for each part of a dataset,
possibly using values external to the part. The input of the algorithm consists
again of a finite set $U$ and an aggregation function $\phi$ that takes a
subset of $U$ and returns a value in some domain $X$.  Compared to
Algorithm~\ref{alg:sac} two different inputs are needed. First, a function must
be defined that that checks whether a given subset $d$ of $U$ is suitable for
computing $\phi(d)$.  We will call this function $\beta: 2^U\to \mathbb{B}$,
where $\mathbb{B}=\{\texttt{True},\texttt{False}\}$.  Typical tests are
checking whether there are sufficient records available, or whether certain
variables have a low enough fraction of missing values. Second, we need a
\emph{collapsing scheme}
\begin{equation}
U\xonto{f}A\xonto{f_1}A_1\xonto{f_2}\cdots\xonto{f_n}A_n.
\label{eq:collapsingsequence}
\end{equation}
A collapsing scheme is a sequence of partitions where each $f_i$ partitions its
domain in $|A_i|$ groups while $f$ partitions $U$ in $|A|$ groups.

Denote with $F_k:A\to A_k$, the function that  accepts a label in $A$ and
returns the corresponding label in $A_k$. In other words, $F_k$ is the
composition $f_k\circ f_{k-1}\circ\cdots \circ f_1$. Similarly we define the
pullback along $F_k$ as $F_k^* = f_1^*\circ f_2^*\circ\cdots\circ f_k^*$.  This
function accepts a set of labels in $A_k$ and returns all the labels in $A$
that are mapped to those labels via the collapsing sequence of
Equation~\eqref{eq:collapsingsequence}.  With this notation we can define the
Algorithm for aggregation with dynamic grouping as follows.

\begin{algorithm}[H]
\caption{Split-Apply-Combine with Collapsing Groups}
\label{alg:saccg}
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{A finite set $U$, an aggregator $\phi: 2^U\to X$, a test function $\beta: 2^U\to \mathbb{B}$,
      and a collapsing sequence $U\xonto{f}A\xonto{f_1}A_1\xonto{f_2}\cdots \xonto{f_n}  A_n$.}

\Output{$R$: the value of $\phi$ for every part of $U$, for which a suitable
collapsing group can be found, as a set of triples $(a,k,x)\in A\times
\underline{n}\times X $ where $\underline{n}=\{0,1,\ldots,n\}$.}

$R = \{\}$\;
\For{$a\in A$}{
  $i=0$ \tcp*{Initiate collapse level}
  $d = f^*(\{a\})$ \tcp*{Get subset of $U$}
  \While{$i<n \land \lnot\beta(d)$}{ \label{line:while}
    $i = i+1$ \tcp*{Increase collapse level} 
    $d = (f^*\circ F_i^*\circ F_i)(a)$ \tcp*{Collapse and get subset}
  }\label{line:endwhile}
  \If{$i<n \lor \beta(d)$}{ \label{line:cond}
    $R = R\cup \{(a, i,\phi(d))\}$\;
  } 
}

@@

In this algorithm the collapsing level $i$ is increased until the test is
passed or the maximum collapsing level $n$ is reached
(Lines~\ref{line:while}-\ref{line:endwhile}) %Since the eventual level of
collapsing is determined dynamically by data %circumstances in $U$, the
algorithm also reports the collapsing level $i$ used.  The condition in
Line~\ref{line:cond} ensures that if no suitable dataset is found after the
whole collapsing sequence has been executed, then no answer is returned. This
means that in contrast with Algorithm~\ref{alg:sac} there is no guarantee that
a value for each member of $A$ will be found. For each member of $A$ where
an aggregate is computed, there is a triple $(a,i,\phi(d))$, where $a\in A$
is the label to which the value $\phi(d)$ pertains, and $i$ is the number of
collapses applied to reach a suitable dataset.

Comparing Algorithms~\ref{alg:sac} and~\ref{alg:saccg}, we see that the
standard split-apply-combine algorithm has worst-case runtime complexity
$O(|A|)$, which is equal to the best case $\Omega(|A|)$. The
split-apply-combine with collapsing groups algorithm has best case
$\Omega(|A|)$ equal to standard split-apply-combine but worst case $O(n|A|)$.
In fact, the best case for Algorithm~\ref{alg:saccg} is achieved by setting
$\beta:d\mapsto\texttt{True}$. It is not hard to see that in this case
Algorithm~\ref{alg:saccg} reduces to Algorithm~\ref{alg:sac}. The worst case is
achieved by setting $\beta: d\mapsto \texttt{False}$.
